---
title: "Introduction to Clinical Data - Answer Key"
author: Nicole Black
date: 07/17/22
---

Welcome to your first experience using TCGA data in R! The first data type we will begin with is clinical data (ex: sex, age, etc.)

###### Set up your working environment

Before running this code, make sure you are in the correct working directory (your Github repository) with getwd() and setwd(). Otherwise you’ll have trouble tracking the data down!

As good general practice to keep things organized, I’d advise to create an "analysis_data" folder in the root of your GitHub repository (your qbio_490_name folder), then set that folder as your working directory ("/PATH/TO/qbio_490_name/analysis_data") whenever you are working with TCGA files.

Go ahead and do that now:

```{r Exercise 1.1}
# cd /PATH/TO/qbio_490_name
# mkdir analysis_data
# pwd

setwd("/Users/nblack/Desktop/QBIO_490/qbio_490_nicole/analysis_data")
```

Specifically for RNotebooks, setting a working directory is a bit funky. Make sure you understand setwd() since this is the command you will use when working in R scripts, but for all notebooks from now on, just add your path into the R Setup chunk as seen below.

```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("/Users/nblack/Desktop/QBIO_490/qbio_490_nicole/analysis_data")) 
```

###### Pulling data from TCGA using TCGAbiolinks

You should have already installed both the BiocManager and TCGAbiolinks packages into R. Go ahead and load those packages in now:

```{r Exercise 2.1}

library(BiocManager)
library(TCGAbiolinks)

```

To get the clinical data files onto your computer, use GDCquery(). 

Hint: you'll need to fill in the "project", "data.category", and "file.type" (=xml) arguments. Look up GDCquery() either in the help pane or online to determine which "project" and "data.category" to use. At the bottom of the help pane there are examples that might be useful as well.

```{r Exercise 2.2}
clinical_query <- GDCquery(project = "TCGA-BRCA",
                           data.category = "Clinical",
                           file.type = "xml")
```

After you have queried the data, download it.

```{r Exercise 2.2}
GDCdownload(clinical_query) # ONLY RUN THIS ONCE! it is very slow
```

Lastly, to get the data into an R data frame, use GDCprepare_clinic().

```{r Exercise 2.3}
clinical <- GDCprepare_clinic(query = clinical_query,
                              clinical.info = "patient")
```

###### Exploring the Data

View the structure of the data as well as the first few lines. What two commands should you use?

1. How many rows are there? How many columns?
    ~ 1174 rows, 114 columns
2. What is the data type of $ gender?
    ~ factor (two levels)
3. What is the age in years of the 5th patient (barcode: TCGA-4H-AAAK)?
    ~ 50
4. How old was the 1st patient at their last follow up (barcode: TCGA-3C-AAAU)? HINT: look at data from two columns
    ~ 65
    
```{r Exercise 3.1}
str(clinical)
head(clinical)
```

View the column and row names. What two commands should you use?

```{r Exercise 3.2}
colnames(clinical)
rownames(clinical)
```

View the vital status column. What two notations can be used to view columns? 

```{r Exercise 3.3}
clinical$vital_status

clinical[ , 7]
```

###### Plotting

Choose two variables and make a scatter plot (recall the plot() function).

THINK: what types of variables can you use in a scatter plot (continuous, discrete, categorical). Look these up if you need help!

For an extra challenge, add a plot title, axes titles, and colors to your plot.

```{r Exercise 4.1}
# continuous and/or discrete

five_ten_yr <- ifelse(clinical$days_to_last_followup > 365 * 10, "Red",
                      ifelse(clinical$days_to_last_followup > 365 * 5, "Purple", "Blue"))

plot(x = clinical$age_at_initial_pathologic_diagnosis,
     y = clinical$days_to_last_followup, 
     main = "Age at Diagnosis vs Days to Last Followup", 
     xlab = "Age at Diagnosis", 
     ylab = "Days to last followup", 
     col = five_ten_yr)
```

Choose two new variables and make a box plot (recall the boxplot() function).

THINK: what types of variables can you use in a box plot?

For an extra challenge, add a plot title, axes titles, and confidence intervals to your plot.

```{r Exercise 4.2}
# one categorical and one discrete/continuous

boxplot(formula = clinical$age_at_initial_pathologic_diagnosis ~ clinical$race_list,
        data = clinical,
        xlab = "Race",
        ylab = "Age at Diagnosis",
        main = "Age at Diagnosis vs Race",
        cex.axis = 0.5)
```
